One-stage point transect distance sampling using iterated integrated nested Laplace approximations

### 2021-11-30 paper.tex
naphon tutorial
update discussion and add unmarked citations

\documentclass{stylefile16/statsoc}
\usepackage[a4paper]{geometry}   % otherwise statsoc class doesn't compile to A4 correctly

%\usepackage{setspace}
%\doublespacing

@@

Concerns about uncertainty propagation can be avoided by using one-stage modelling approaches.  One-stage analyses of distance sampling data have tended to be Bayesian and use Markov chain Monte Carlo (MCMC) methods for inference, usually requiring data augmentation to model unobserved individuals or groups
% problem with this Royle reference here.  Not sure what I was intending to cite.
\citep{royle_HierarchicalModelSpatial_2008, schmidt_using_2012}.  \citet{oedekoven_bayesian_2014} present a one-stage model that avoids data augmentation and incorporate model uncertainty using reversible jump MCMC.

@@

The model fits 40 to 60 seconds on a laptop with 10 Gb RAM and a 2.50 GHz CPU when INLA is run using the `empirical Bayes' option which does not incorporate uncertainty in hyper parameters.  The model fits in minutes in full INLA mode including hyper parameter uncertainty.  We suggest using the `empirical Bayes' mode for model development and full INLA on a smaller subset of candidate models.

@@

The SD values also reflect a modelling \textit{assumption} but not necessarily variability due to varying survey effort.  The SD is linked to the predicted mean by the assumed log-Gaussian relationship which means the mean and variance are directly connected, with higher mean values necessitating higher SD values.  This means that the SD map will produce a pattern similar to the mean intensity map, as is the case here (Figures 2A and 2B).  This relationship is the result of a modelling assumption and not necessarily reflective of the detectability of animals or survey effort.  One might expect uncertainty to be lower in areas with greater survey effort, however for the \akepa{} data the opposite is the case when interpreting the SD map.

@@

This paper presents a new approach to analysing point transect distance sampling data that incorporates several innovative methods and considers the bulk of an applied statistician's workflow of model specification, inference procedure, model evaluation and communication of results.  Although any of these areas on its own would be substantive enough to be the main focus of a paper, our brief presentation of issues in sequence along with our chosen solutions will be useful to statisticians who face similar questions and data structures.

There are several natural extensions to the model.  For this analysis we only considered a single survey year.  Spatio-temporal extensions to this are possible, for example by considering a space-time interaction models.  \texttt{R-INLA} has the ability to fit space-time effects that can be represented as a Kronecker product of the relevant precision matrices \citep{blangiardo_spatial_2013, yuan_point_2017}.  An example of this type of space-time interaction effect is an interaction between the spatial SPDE with an auto-regressive temporal process.  \cite{blangiardo_spatial_2013} show how this can be done with numerous examples and code snippets.

Another natural extension is to consider more complicated observation processes.  There is the potential to include explanatory covariates on detection function parameters.  For example, we did not make use of the detection type (audio or visual) in our analysis, although this was recorded and detectability may vary depending on whether a bird is seen or heard.  Other factors could be considered, such as weather conditions, observer expertise, animal behaviour and morphological traits.  Any additional parameters to the non-linear model components will be estimated using the outer-optimisation step of iterated INLA.  [[Can I speculate here about how iterated INLA might perform as parameters are added to non-linear components?  Is, higher-dimensional optimisation slower/infeasible/unstable at a certain point?]]

The iterated INLA inference procedure is a general approach that could be applied to other non-linear model components that arise in the area of spatial ecology.  We note two promising areas in which our flexible inference procedure may be useful:

\begin{itemize}
	\item \textbf{Selectivity analysis} In surveys of the marine environment, the size, length and weight of fish affects the likelihood of being caught in a trawl net.  Thus any analysis of these data must account for this.  The probability of capture is typically modelled as a logistic function with parameters that depend on morphological traits \citep{herrmann_understanding_2016, madsen_selectivity_2007, galbraith_demersal_1994}, which could be estimated jointly with the spatial distribution within our inference framework.
	\item \textbf{Functional responses} In the area of understanding species-habitat or species-species interactions, the concept of a `functional response' captures the idea that the abundance of a resource or risk will affect the response to it \citep{holling_some_1959}.  These varying responses can be described mathematically using parametric equations that have the desired properties, e.g. a horizontal asymptote so that beyond a certain threshold no extra amount of a resource will affect the response to it.  These types of responses are typically not considered in species distribution models but our framework allows them to be estimated alongside the more common linear or quadratic fixed effects. This flexibility in specifying model components could also be more widely applicable in areas such as public health and spatial epidemiology, where responses to conditions may be non-linear.

\end{itemize}
We note that many of the themes addressed in this paper apply beyond the field of spatial ecology and will be of interest to the spatial statistics community in general, particularly for applications with complex observation models of latent processes and where spatial predictions are a key output of the analysis

In our analysis we only briefly touch upon the problem of communicating uncertainty in maps and highlight the excursions methods as a new and positive addition to this area.  In our example, the threshold was chosen to illustrate the method but we believe that discussions with relevant stakeholders will lead to agreement on a (possibly large) set of thresholds that are relevant given the aims and context of the analysis.  This requires the input of both the relevant domain experts and statisticians so stakeholders understand the consequences of the thresholds and uncertainty levels they wish to consider.  We also note the potential for excursions to be used to define the notion of occupancy (and thus map occupied areas, with a given probability of error).

We end with a note on the similarity between what we have done and existing two stage approach.  The one-stage thinned point process perspective can seem fundamentally different to a two-stage GAM approach fitted to count data.  It can be tempting to think that, because the point process model does not require us to bin the data into counts, we can avoid some loss of information.  However, we also made the same assumption of constant intensity within each transect as in more traditional distance sampling approaches.  This leads to the same loss of information since we do not use the exact location of points within each transect, except to account for detectability.

In summary, we have presented a novel framework for analysing point transect distance sampling data that introduces new approaches to model specification, inference procedure and communication of results.  These developments will be of interest to those with similar data structures in spatial ecology as well as researchers in other areas who we hope will explore the full flexibility and generality of these developments in their own work.

% spp has potential to provide general framework? [Janine comment]
% Mention modelling on sphere is possible without projecting on to R^2?

>>>>

\documentclass{stylefile16/statsoc}
\usepackage[a4paper]{geometry}   % otherwise statsoc class doesn't compile to A4 correctly
%\usepackage{setspace}
%\doublespacing

@@

Concerns about uncertainty propagation can be avoided by using one-stage modelling approaches.  A popular one-stage maximum-likelihood method, introduced by \cite{royle_ModelingAbundanceEffects_2004} and implemented in the \texttt{R} package \texttt{unmarked} \citep{fiske_UnmarkedPackageFitting_2011}, is based on reformulating the model as a multinomial likelihood.  This depends choosing discrete distance classes and binning the data into counts defined on discrete spatial units.  Given these discrete units, the likelihood can be written in a multinomial form and, by marginalising out site specific abundance, a Poisson likelihood that can be optimised to achieve one-stage inference.  This approach rests heavily on the discretisation of both the distance data and the spatial location data.  Below we present a point process perspective that does not require this discretisation step in either the distance or spatial location data.  The \texttt{unmarked} package also does not implement spatially structured random effects, something that we use here and is also popular in the two-stage approach implemented using \texttt{mgcv}.

Bayesian one-stage approaches have tended to use Markov chain Monte Carlo (MCMC) methods for inference along with data augmentation to model unobserved individuals or groups \citep{schmidt_using_2012}.  \citet{oedekoven_bayesian_2014} present a one-stage model that avoids data augmentation by specifying a combined likelihood of the detection and spatially-explicit count models and incorporated model uncertainty using reversible jump MCMC.

@@

The model fits in roughly 40 to 60 seconds on a laptop with 10 Gb RAM and a 2.50 GHz CPU when INLA is run using the `empirical Bayes' option which does not incorporate uncertainty in hyper parameters.  The model fits in a few minutes in the full INLA mode, including hyper parameter uncertainty.  We suggest using the `empirical Bayes' mode for model development and full INLA on a smaller subset of candidate models.

@@

The SD values also reflect a modelling assumption rather than variability due to, say, varying survey effort.  The SD is linked to the predicted mean by the assumed log-Gaussian relationship.  If X is normally distributed with mean $\mu$ and variance $\sigma^2$ then $Z = \exp(X)$ has a mean-variance relationship given by $\text{Var}(Z) = \exp(\sigma^2 - 1)\mathbb{E}(Z)^2$. Therefore the SD map will generally show a pattern similar to the mean intensity map (Figures 2A and 2B).  Since this relationship is the result of a modelling assumption and not necessarily reflective of the detectability of animals or survey effort, one should take care when interpreting the SD map.  It should be noted however, that this modelling assumption affects uncertainty quantification for all outputs based on the posterior intensity field, not just the SD map.

@@

We have presented a new approach to analysing point transect distance sampling data in a Bayesian setting.  We chose to do this in a relatively simplified setting, considering a single year of data, with no covariate effects and using a GRF to explain spatial variation in density.  There are several natural extensions to the model.  Spatio-temporal models can also be fitted using \texttt{R-INLA}, which would allow multiple years of data to be modelled jointly, as was the case in \citep{camp_dsm_2020}.  Another extension is to develop the observation process model by considering other detection functions or including covariates, for example.  Factors such as weather conditions, observer expertise, animal behaviour and morphological traits are likely to affect detectability.  Future research should investigate the degree of complexity that can be well-approximated by the iterated INLA approach.

The iterated INLA inference procedure has the potential to be applied to other non-linear model components that arise spatial statistics.  We note two promising areas within the field of spatial ecology that highlight this potential:

\begin{itemize}
	\item \textbf{Selectivity analysis} In surveys of the marine environment, the size, length and weight of fish affects the likelihood of being caught in a trawl net.  The probability of capture is typically modelled as a logistic function with parameters that depend on morphological traits \citep{herrmann_understanding_2016, madsen_selectivity_2007, galbraith_demersal_1994}.  This function is analogous to the detection function in distance sampling models.  Our framework could allow, for example, the joint modelling of selectivity and spatial data.
	\item \textbf{Functional responses} The concept of a `functional response' captures the idea that the overall abundance of a resource or risk will affect a species' response to it \citep{holling_some_1959}.  These varying responses can be described mathematically using parametric equations that have the desired properties, e.g. an increasing function with a horizontal asymptote, to capture the notion of diminishing returns.  Such non-linear functions could be incorporated within our framework.  This flexibility in specifying model components could also be more widely applicable in areas such as public health and spatial epidemiology, where responses to conditions may suitably be modelled using non-linear relationships.
\end{itemize}
There are no doubt many more application areas that could make use of the additional flexibility that comes with allowing non-linear predictors.

The popularity of using INLA for spatial modelling has grown, in large part, due to the computational efficiency of the approach.  However, this efficiency comes at the cost of reduced flexibility in model specification over, say, a more general probabilistic programming languages such as \texttt{JAGS} \citep{plummer_jags_2017} or \texttt{Stan} \citep{stan_Rstan_2020}.  The iterated INLA approach allows a wider class of models to be considered whilst retaining much of the computational benefits of INLA.  In addition to computational efficiency, there are some features of INLA that are not generally available in other packages, such as: (i) the ability to model using GRFs directly on the sphere, avoiding projecting to $\mathbb{R}^2$ \citep{lindgren_explicit_2011}; (ii) GRFs that account for complex barriers such as coastlines and islands, to avoid inappropriately smoothing across these \cite{bakka_NonstationaryGaussianModels_2019}; and (iii) support for joint likelihood models, allowing modelling of, for example, marked point processes \citep{illian_FittingComplexEcological_2013}, and combining multiple data sources.  These approaches, and many others we have not mentioned, that are implemented in INLA, can now readily be incorporated into distance sampling analyses, as well as broader spatial ecology and spatial statistics applications.
